---
title: "Chudley preprocessing"
output: html_document
---

*Packages*
```{r}
setwd("~/Chudley_NDN/")
library(methylumi)
library(gplots)
library(marray)
library(lumi)
library(lattice)
library(RColorBrewer)
library(knitr)
library(xtable)

```

*Loading files*
```{r}
sampleFile <- ("~/Chudley_NDN/chudley_NDN_metadata_new.csv") #text file with phenotype variables: sex, age, etc.
allFile <- ("~/Chudley_NDN/GS_Jan_2016/chudley-alldata.txt") 
betaFile <- ("~/Chudley_NDN/GS_Jan_2016/chudley-avgbeta.txt")
qcFile <- ("~/Chudley_NDN/GS_Jan_2016/chudley-qcfile.txt") 

file.exists(allFile)
file.exists(betaFile)
file.exists(qcFile)
file.exists(sampleFile)
```

*Methylumi files*
```{r}
chudley.raw <- lumiMethyR(allFile)
chudley.2.raw <- methylumiR(betaFile, qcfile = qcFile)

dim(chudley.raw) # probes = 485,577, n = 53
dim(chudley.2.raw) # probes = 485,577, n = 53
```

*Add metadata*
```{r}
sampleInfo <- read.csv(sampleFile, row.names=1, header = T,stringsAsFactors=F)
stopifnot(all(sampleNames(chudley.raw)%in%rownames(sampleInfo))) #TRUE (same order with the same sample names?)

head(sampleInfo <- sampleInfo[sampleNames(chudley.raw),])
str(pData(chudley.raw) <- sampleInfo)
View(sampleInfo)

chudley.raw <- chudley.raw[, order(chudley.raw$Sentrix_ID)]
chudley.2.raw <- chudley.2.raw[, sampleNames(chudley.raw)]
sampleInfo <- sampleInfo[order(sampleInfo$Sentrix_ID),]

```

*Save raw files*
```{r}
#save(chudley.raw, file ='chudley(raw).RData')
sampleNames(chudley.2.raw)
sampleNames(chudley.raw)
sampleNames(chudley.2.raw) <- rownames(sampleInfo)
#save(chudley.2.raw, file ='chudley.2(raw).RData')

#from now on will be using chudley_wc object, these are our "working projects". Will save other objects as we go...
load("~/Chudley_NDN/chudley(raw).RData")
load("chudley.2(raw).RData")

chudley_wc <- chudley.raw
dim(chudley_wc) #probes = 485577, samples = 53
chudley.2_wc <- chudley.2.raw
dim(chudley.2_wc) #probes = 485577, samples = 53
```

```{r}
par(cex = 0.7)
par(mfrow=c(1,1))
plotSampleRelation(chudley_wc, method = "mds", col = as.factor(chudley_wc$Sentrix_ID), addLegend=F)
#WH014_1 looks like an outlier
#legend ("top", legend = levels (factor (chudley_wc$Sentrix_ID)), fill = 1:nlevels (factor (chudley_wc$Sentrix_ID)), horiz = F)

plotSampleRelation(chudley_wc, method = "mds", col = as.factor(chudley_wc$Sex), addLegend=F) 
#legend ("top", legend = levels (factor (chudley_wc$Sex)), fill = 1:nlevels (factor (chudley_wc$Sex)))
```

*Add magdas annotation*
```{r}
magprice <- read.delim("~/Rscripts/Price_platform_only.txt", row.names=1)
colnames(magprice) <- paste(colnames(magprice), "price", sep="_")
fData(chudley_wc) <- cbind(fData(chudley_wc), magprice)
dim(chudley_wc) #485512

fdat <- fData(chudley_wc)
```


*Remove SNP & XY probes*
```{r}
chudley.rs <- chudley_wc[substring(featureNames(chudley_wc), 1, 2) == "rs", ] 
dim(chudley.rs) # SNP probes = 65, n = 53
snp_probes65 <- rownames(chudley.rs)
save(snp_probes65, file = "chudley_snp_probes65.RData")

par(cex = 0.7)
plotSampleRelation(chudley.rs, method = "cluster")
#WD310B/D are replicates

chudley_wc <- chudley_wc[substring(featureNames(chudley_wc),1,2) != "rs", ]

dim(chudley_wc) # probes = 485,512, n = 53
#save(chudley.rs, file = "chudley(onlySNPprobes).RData")

# Removing XY Probes (targets the X or Y) 
chudley.xy <- chudley_wc[fData(chudley_wc)$CHR%in%c("X","Y"),]
dim (chudley.xy) # 11648, 53

XYprobes <- rownames(chudley.xy)
save(XYprobes, file = "chudley_XYprobes.RData")
chudley_wc<- chudley_wc[!fData(chudley_wc)$CHR%in%c("X","Y"),]
dim(chudley_wc) # 473864, 53
```

*Sex sanity check*
```{r}
XYcor <- cor(betas(chudley.xy), use = "pairwise.complete.obs")

sexCol <- chudley_wc$Sex
sexCol <- gsub("M", "black", gsub("F", "grey", sexCol))
library(RColorBrewer)
library(gplots)
BuPu <- brewer.pal(9, "BuPu")
plot.new()
heatmap.2(XYcor,main = "Correlation of samples using XY probes (n = 11,648",
          trace = "none", col = BuPu, dendrogram = "both",
          ColSideColors = sexCol, cexCol = 0.9,
          keysize = 1)
legend("top", bty = "n", title = "top bar colours", c("Male", "Female"), cex = 1,
       fill = c("black", "grey"), ncol = 2)
#WH004 is definitely a male, not a female
pData(chudley_wc)$Sex[grep("WH004",rownames(pData(chudley_wc)))] <- "M"
#fixed

```


*Checking which samples to remove
```{r}
pvals(chudley.raw)
avgPval <- colMeans(pvals(chudley.raw))
barplot(avgPval,ylab="Average P-Value",width=800) # barplot of samples average pvals

boxplot(estimateIntensity(exprs(chudley.raw)))
sample.Pval <- detectionCall(chudley.raw, Th=0.01, type="sample")
which((sample.Pval>485577 -(485577*.05))==FALSE) #WH026, WD107_1_Rep1, WH014_1 (samples with 1% of site det p>0.05 == WD294_2,WH008A_1, WH270 also)
mldatM.samp <- chudley.raw[,(sample.Pval>485577 -(485577*.05))] #3 samples with  >5% sites with det p>0.05 removed
485577 - (485577*.01)
dim(mldatM.samp) 

plotSampleRelation(mldatM.samp, method = "cluster", horiz = T)
length(which(pData((mldatM.samp))$Group == "Control"))
heatmap.2(cor(mldatM.samp), trace="none")

density(mldatM.samp, legend= "none")

avgPval <- colMeans(pvals(chudley.raw))
barplot(avgPval,ylab="Average P-Value", ylim=c(0,0.006), las=2)


boxplot(estimateIntensity(chudley.raw))
density(estimateIntensity(chudley.raw), xlab="log2(CpG-site Intensity)")
hist(chudley.raw)


```


```{r}
# Removing bad probes
all(featureNames(chudley_wc)%in%featureNames(chudley.2_wc)) #true
all(sampleNames(chudley_wc)%in%sampleNames(chudley.2_wc))  #true
avgbeta <- betas(chudley.2_wc)[featureNames(chudley_wc),sampleNames(chudley_wc)]
badAvgbeta <- is.na(avgbeta)
dotchart(colSums(badAvgbeta), cex=0.4, main = paste("Bad Avg Beta: probes = 473,864"))
nbadAvgbeta <- print(sum(rowSums(badAvgbeta)>=5)) # 762 (758 when removing WH014_1 less )

badDetP <- detection(chudley_wc)>0.01
nbadDetP <- print(sum(rowSums(badDetP)>=5)) # 17451 (1749 come from WH014_1 ~10%)
dotchart(colSums(badDetP),main=paste("Bad Det P, probes = 473,684"), cex = 0.4)
      #worst replicates = WH004_rep2, WD107_1_rep1, WD310B

badProbes<-rownames(badAvgbeta)[rowSums(badAvgbeta)>=5|rowSums(badDetP)>=5]
length(badProbes) # 18064

chudley_wc <- chudley_wc[!featureNames(chudley_wc)%in%badProbes] 

dim(chudley_wc) # 455800, 53
```


```{r}
# Removing polymorphic CpGs probes #19078
snp_index <- which(is.na(fData(chudley_wc)$n_SNPCpG_price))
chudley_wc<- chudley_wc[snp_index, ] 
dim(chudley_wc) # 436722, 36
455800 - 436722

colnames(fdat)
# Removing multiple XY binder probes #10446
xy_hit_index <- which(fData(chudley_wc)$XY_Hits_price == "XY_NO") 
chudley_wc<- chudley_wc[xy_hit_index, ]
dim(chudley_wc) # 426276 , 53
436722-426276
#485577 - 426276 =    59301 removed ### 
```


*Removing bad samples*
```{r}
load("~/Chudley_NDN/chudley.2(raw).RData")
chudley_wc.noout <- chudley.raw 
chudley.2_wc.noout <- chudley.2.raw
chudley_wc.noout <- chudley_wc.noout[,!(colnames(chudley_wc.noout) %in% c("WH014_1", "WH026", "WD107_1_Rep1"))]
chudley.2_wc.noout <- chudley.2_wc.noout[,!(sampleNames(chudley.2_wc.noout) %in% 
                                              c("WH014_1", "WH026", "WD107_1_Rep1"))]
dim(chudley_wc.noout)
dim(chudley.2_wc.noout)

chudley_wc.noout <- chudley_wc.noout[substring(featureNames(chudley_wc.noout),1,2) != "rs", ]
dim(chudley_wc.noout)
magprice <- read.delim("~/Rscripts/Price_platform_only.txt", row.names=1)
colnames(magprice) <- paste(colnames(magprice), "price", sep="_")
fData(chudley_wc.noout) <- cbind(fData(chudley_wc.noout), magprice)
dim(chudley_wc.noout) #485512

chudley_wc.noout <- chudley_wc.noout[!fData(chudley_wc.noout)$CHR%in%c("X","Y"),]
pData(chudley_wc.noout)$Sex[grep("WH004",rownames(pData(chudley_wc.noout)))] <- "M"
dim(chudley_wc.noout) #473,864


# Removing bad probes
all(featureNames(chudley_wc.noout)%in%featureNames(chudley.2_wc.noout)) #true
all(sampleNames(chudley_wc.noout)%in%sampleNames(chudley.2_wc.noout))  #true
avgbeta.noout <- betas(chudley.2_wc.noout)[featureNames(chudley_wc.noout),sampleNames(chudley_wc.noout)]
badAvgbeta.noout <- is.na(avgbeta.noout)
dotchart(colSums(badAvgbeta.noout), cex=0.4, main = paste("Bad Avg Beta: probes = 473,864"))
nbadAvgbeta.noout <- print(sum(rowSums(badAvgbeta.noout)>=5)) # 726

badDetP.noout <- detection(chudley_wc.noout)>0.01
nbadDetP.noout <- print(sum(rowSums(badDetP.noout)>=5)) # 11864 (~10% came from WH014_1)
dotchart(colSums(badDetP.noout),main=paste("Bad Det P, probes = 473,684"), cex = 0.4)
      #worst replicates = WH004_rep2, WD107_1_rep1, WD310B

badProbes.noout<-rownames(badAvgbeta.noout)[rowSums(badAvgbeta.noout)>=5|rowSums(badDetP.noout)>=5]
length(badProbes.noout) # 12,459

chudley_wc.noout <- chudley_wc.noout[!featureNames(chudley_wc.noout)%in%badProbes.noout] 

dim(chudley_wc.noout) # 461,405, 50

# Removing polymorphic CpGs probes #19413
snp_index.noout <- which(is.na(fData(chudley_wc.noout)$n_SNPCpG_price))
chudley_wc.noout<- chudley_wc.noout[snp_index.noout, ] 
dim(chudley_wc.noout) #  442028, 48
461405 - 442028 = 19337


# Removing multiple XY binder probes #10455
xy_hit_index.noout <- which(fData(chudley_wc.noout)$XY_Hits_price == "XY_NO") 
chudley_wc.noout <- chudley_wc.noout[xy_hit_index.noout, ]
dim(chudley_wc.noout) # 431,544  , 48
442028-431544 = 10484
#485577 - 431544 =    54033 removed ### 

save(chudley_wc.noout, file = "chudley.noout(preprocessed).RData")
```


*** NORMALIZATION ***

```{r}
load("~/Chudley_NDN/chudley.noout(preprocessed).RData")
chudley_norm <- chudley_wc.noout
heatmap.2(cor(exprs(chudley_wc.noout)), trace="none")
quantile(cor(exprs(chudley_wc.noout)), probs = c(0,0.25,0.50,0.75,0.95))

dim(chudley_norm) # 431544, 50
chipCol <- rainbow(nlevels(factor(chudley_norm$Sentrix_ID)))[as.numeric(factor(chudley_norm$Sentrix_ID))]
density(chudley_norm, col = chipCol, lty = 1, addLegend=F,
        main = "chudley_noout after probe filtering, before col.corr and norm. - M-values by batch")

chudley_norm.q <- lumiMethyN(chudley_norm,method="quantile")

```

*Subset-quartile within array (SWAN) normalization*
```{r}
target <- fData(QCdata(chudley.2_wc.noout))$TargetID =="NEGATIVE"
grn <- methylated(QCdata(chudley.2_wc.noout))[target,]
red <- unmethylated(QCdata(chudley.2_wc.noout))[target,]
bkg <- rowMeans(cbind(colMeans(grn),colMeans(red)))

#sanity check
all(names(bkg)%in%sampleNames(chudley_norm.q)) #TRUE
bkg <- bkg[sampleNames(chudley_norm.q)]


#Let's load the required script for SWAN, and then run it.
source("~/Rscripts/SWAN scripts.R")
load("~/Rscripts/nCpG all.RData")

all(featureNames(chudley_norm.q)%in%names(nCpG.all)) #TRUE
fData(chudley_norm.q)$nCpG <- nCpG.all[featureNames(chudley_norm.q)]

chudley_swan <- preprocessMSWAN(chudley_norm.q,bg=bkg)

sum(is.na(exprs(chudley_swan))) # should be 0
sum(is.na(unmethylated(chudley_swan))) # should be 0
sum(is.na(methylated(chudley_swan))) # should be 0


#Get the M and beta value outputs from CAM_wc.swan:
M.chudley.swan <-exprs(chudley_swan)
# write.table(M.chudley.swan,file='chudley.swan_M.txt')

betas.chudley.swan <- m2beta(chudley_swan)
# write.table(betas.chudley.swan,file='chudley.swan_B.txt')

# save(chudley_swan,file = "chudley.swan.RData")
```

*Peak-based correction*
```{r}
chudley_pbc <- chudley_norm
modifiedM <- peakBasedCorrection(chudley_norm)
exprs(chudley_pbc) <- as.matrix(modifiedM[featureNames(chudley_pbc), sampleNames(chudley_pbc)])
dim(chudley_pbc)

```



*Plot beta values before/after normalization*
```{r} 
type <- fData(chudley_norm)$INFINIUM_DESIGN_TYPE

before.betas <- betas(chudley_norm)
dim(before.betas)

after.betas.SWAN <- betas(chudley_swan)
dim(after.betas.SWAN)

after.betas.PBC <- betas(chudley_pbc)

plot.new()
par(mfrow = c(1, 1))
plot(c(0, 1) , c(0, 10), type = "n", xlab = "Beta values", ylab = "Density")
lines(density(na.omit(before.betas)[type=="I",]),col="blue",lty=1, lwd =3)
lines(density(na.omit(before.betas)[type=="II",]),col="blue",lty=2, lwd=3)
lines(density(na.omit(after.betas.SWAN)[type=="I",]),col="firebrick2",lty=1, lwd = 3)
lines(density(na.omit(after.betas.SWAN)[type=="II",]),col="firebrick2",lty=2, lwd=3)
lines(density(na.omit(after.betas.PBC)[type=="I",]),col="green",lty=1, lwd = 3)
lines(density(na.omit(after.betas.PBC)[type=="II",]),col="green",lty=2, lwd=3)
legend("top", inset=.05, cex=0.8, c("Before Type 1","Before Type 2", "SWAN Type 1", "SWAN Type 2"), col=c("blue","blue","firebrick2","firebrick2"), lwd=3, lty=c(1,2,1,2), horiz=FALSE)
```
#peak based correction looks much better

*Replicates*
```{r}
sampleNames(chudley_norm)
# WD107_1_Rep2 <- Rep1 removed because of bad probes
# WH004A_Rep1 & WH004_2_Rep2
# WD310B & WD310D

cor(exprs(chudley_norm[, grep("WH004", sampleNames(chudley_norm))])) #0.9947
cor(exprs(chudley_norm[, grep("WD310", sampleNames(chudley_norm))])) #0.9943


WH004A.raw <- chudley_norm[, grep("WH004", sampleNames(chudley_norm))]
WH004A.pbc <- chudley_pbc[, grep("WH004", sampleNames(chudley_pbc))]
WH004A.swan <- chudley_swan[, grep("WH004", sampleNames(chudley_swan))]

WD310.raw <- chudley_norm[, grep("WD310", sampleNames(chudley_norm))]
WD310.pbc <- chudley_pbc[, grep("WD310", sampleNames(chudley_pbc))]
WD310.swan <- chudley_swan[, grep("WD310", sampleNames(chudley_swan))]


plot.new()
par(mfrow = c(2,3))
density(WH004A.raw,legend="bottom", lty = 1, xlim = c(-6, 6), main = "WH004 quantile")
density(WH004A.pbc,legend="bottom", lty = 1, xlim = c(-6, 6), main = "WH004 pbc")
density(WH004A.swan,legend="bottom", lty = 1, xlim = c(-6, 6), main = "WH004 swan")

density(WD310.raw,legend="bottom", lty = 1, xlim = c(-6, 6), main = "WD310 raw")
density(WD310.pbc,legend="bottom", lty = 1, xlim = c(-6, 6), main = "WD310 pbc")
density(WD310.swan,legend="bottom", lty = 1, xlim = c(-6, 6), main = "WD310 swan")
par(mfrow = c(1, 1))


#removal of replicates -- worst replicates from earlier = WH004_2_Rep2 & WD310B

chudley_clean <- chudley_swan[,!(colnames(chudley_swan) %in% c("WH004_2_Rep2", "WD310B"))]
dim(chudley_clean) #431544, 48 
#save(chudley_clean, file = "chudley_preprocessed_norm.RData")

density(chudley_clean, legend="none")

plot.new()
par(mfrow = c(3,4))
for(i in 1:48){
  tryCatch(density(chudley_clean[,i], xlim = c(-6, 6), ylim = c(0,0.5), legend="none"), error=function(e) NULL)
  print(i)
}

length(which(pData(chudley_clean)$Group == "Control")) #24
length(which(pData(chudley_clean)$Group == "FASD")) #24
sampleNames(chudley_clean)


```






