---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
library(mediation)
setwd("path/to/data")
```

*Function to extract mediation summary data*
```{r extract_mediation_summary}
extract_mediation_summary <- function (x) { 

  clp <- 100 * x$conf.level
  isLinear.y <- ((class(x$model.y)[1] %in% c("lm", "rq")) || 
                   (inherits(x$model.y, "glm") && x$model.y$family$family == 
                      "gaussian" && x$model.y$family$link == "identity") || 
                   (inherits(x$model.y, "survreg") && x$model.y$dist == 
                      "gaussian"))

  printone <- !x$INT && isLinear.y

  if (printone) {

    smat <- c(x$d1, x$d1.ci, x$d1.p)
    smat <- rbind(smat, c(x$z0, x$z0.ci, x$z0.p))
    smat <- rbind(smat, c(x$tau.coef, x$tau.ci, x$tau.p))
    smat <- rbind(smat, c(x$n0, x$n0.ci, x$n0.p))

    rownames(smat) <- c("ACME", "ADE", "Total Effect", "Prop. Mediated")

  } else {
    smat <- c(x$d0, x$d0.ci, x$d0.p)
    smat <- rbind(smat, c(x$d1, x$d1.ci, x$d1.p))
    smat <- rbind(smat, c(x$z0, x$z0.ci, x$z0.p))
    smat <- rbind(smat, c(x$z1, x$z1.ci, x$z1.p))
    smat <- rbind(smat, c(x$tau.coef, x$tau.ci, x$tau.p))
    smat <- rbind(smat, c(x$n0, x$n0.ci, x$n0.p))
    smat <- rbind(smat, c(x$n1, x$n1.ci, x$n1.p))
    smat <- rbind(smat, c(x$d.avg, x$d.avg.ci, x$d.avg.p))
    smat <- rbind(smat, c(x$z.avg, x$z.avg.ci, x$z.avg.p))
    smat <- rbind(smat, c(x$n.avg, x$n.avg.ci, x$n.avg.p))

    rownames(smat) <- c("ACME (control)", "ACME (treated)", 
                        "ADE (control)", "ADE (treated)", "Total Effect", 
                        "Prop. Mediated (control)", "Prop. Mediated (treated)", 
                        "ACME (average)", "ADE (average)", "Prop. Mediated (average)")

  }

  colnames(smat) <- c("Estimate", paste(clp, "% CI Lower", sep = ""), 
                      paste(clp, "% CI Upper", sep = ""), "p-value")
  smat

}
```

*Loading mediation sites from ALSPAC*
```{r alspac sites}
alspac <- gdata::read.xls("/OneMind-FFCWS/Mediation/Mediation_sites_all_2021-06-04.xls", sheet=1)
head(alspac)
dim(alspac)
alspacLoci <- unique(as.character(alspac$CpG))
alspacLoci <- alspacLoci[!alspacLoci ==""]
length(alspacLoci) #67 loci
rm(alspac)
```
Notes: for simplicity sake, I am going for the shotgun approach and will filter CpGs by adversity after the fact. 

*Loading in FFCWS phenotype data* - from Jonah's call_slcma.rmd (OneMind-FFCWS/Results/5-20-2021/call_slcma.rmd)
```{r load in ffcw data}
library(haven)
pd <- readRDS("../data/analytic/pd_formatted.rds") %>% as.data.table
pd <- pd[cut == ""]
pd[, cm1bsex := sex]

#Filter technical replicates
setkey(pd, probe_fail_pct)
pd <- pd[childteen == "C"] 
pd <- pd[!duplicated(idnum)] #pulling out technical replicates
setkey(pd, idnum)

pd <- pd[, .(MethID, idnum, cm1bsex, childteen, ck6ethrace, leukocyte)][, idnum := .(as.character(idnum))]

#Tmp contains additional covariate data and exposures
tmp <- merge(
  fread("data/input/SLCMA.csv")[, idnum := .(as.character(idnum))],
  read_stata("data/input/SLCMA_II.dta") %>% as.data.table %>% .[, idnum := .(as.character(idnum))],
  by = "idnum"
)

tmp[, colnames(tmp)[endsWith(colnames(tmp), "a")] := NULL] #Alternative parenting measure filtered out 
pd <- merge(pd, tmp, by = "idnum")
rm(tmp)
```

*Add race and location covariates* - from Jonah's call_slcma.rmd
```{r race and 3-city location}
load("../general_data/mfcrace.rdata")
race <- data.table(mfcrace)[, ck6ethrace := NULL] ; rm(mfcrace)
pd <- merge(pd, race, by = "idnum")

#----Impute race for missing cases first mother then father
pd[as.numeric(ck6ethrace) < 0, ck6ethrace := cm1ethrace]
pd[as.numeric(ck6ethrace) < 0, ck6ethrace := cf1ethrace]

#----
load("../general_data/SAND brain vol.rdata")
city <- data.table(SAND_brain_vol) ; rm(SAND_brain_vol)

#---- Provide city variable
pd[, city := .(idnum %in% city[, idnum])]

rm(city)
rm(race)
```

*Format basic covariates* - from Jonah's call_slcma.rmd
```{r formatting covariate classes}
covars <- c(childsex = "cm1bsex", prevpregs = "priorkids", medu = "medu0", mage = "mage0", bwt = "bwgfinal", spreg = "smoke0", cell = "leukocyte", city = "city", race = "ck6ethrace")
tofactor <- c("cm1bsex", "medu0", "smoke0", "city", "ck6ethrace")
tonumeric <- c("priorkids", "mage0", "bwgfinal", "leukocyte")

pd[, (tofactor) := lapply(.SD, factor), .SDcols = tofactor]
pd[, (tonumeric) := lapply(.SD, as.numeric), .SDcols = tonumeric]
pd[, c("oneadult15","dep15") := NULL]

#Set "ID" column as idnum
pd[, ID := as.character(idnum)]
```

*List exposures and set as factor* - adapted from Jonah's call_slcma.rmd
```{r time ages}
exposures <- c("oneadult", "dep", "par", "fh", "faminst")
pattern <- paste0("^", exposures) %>% paste(collapse = "|")
exfactor <- grep(pattern, colnames(pd), value = T)

pd[, (exfactor) := lapply(.SD, as.factor), .SDcols = exfactor]

#making an "ever-exposed" version of those variables
for(i in exposures){
  temp <- pd[,grep(i, colnames(pd))]
  temp$ever <- ifelse(rowSums(temp, na.rm=T) >0, 1, 
                      ifelse(rowSums(is.na(temp))>0, NA, 0)) 
    #any exposure = exposed, unexposed = all unexposed without any missingness
  temp$ever <- as.factor(temp$ever)
  colnames(temp)[ncol(temp)] <- paste0(i, ".ever")
  pd <- cbind(pd, temp[,ncol(temp)]) #adding to the pd dataframe
    
}

```

*Adding depression as outcome* - Jonah, please add here
```{r}
#add a pd column called depression, which is the CBCL at age 15 
#also wonder whether you might do this for CBCL at age 9, but probably too close to the DNAm age

```

*Reading in DNAm data - age 9* - if merging 450K and EPIC, would include a covariate for array above 
```{r read in DNAm data}
beta <- readRDS("../general_data/beta.rds")
beta <- beta[, match(pd[, MethID], colnames(beta))]

#selecting only CpGs from ALSPAC analyses 
beta <- beta[rownames(beta) %in% alspacLoci, ] 
beta <- t(beta) #mediation script is written to take CpGs as column names

```


*Function to run mediation analyses*
```{r runMediation function}
#dnam has rownames as ID and colnames as cpg name; these should be the celltype adjusted m-value
#pheno has an ID column, depression, and exposure columns, as well as covariates
#exposure is the name of the exposure column
#pheno = pd
#dnam = beta
#covars = covars
#covarstmp <- covars[!covars %in% "ck6ethrace"]
# 

runMediation <- function(pheno, dnam, exposure, loci){
  #error messages
  if(exists("pheno") == F){
    stop("pheno not found 
         Please include phenotype file with ID, exposure, depression, and following covariates:
         cm1bsex; priorkids; medu0; mage0; bwgfinal; smoke0; leukocyte; city")}
  if(sum(colnames(pheno) %in% c("ID", exposure, "depression","cm1bsex", "priorkids", 
                                "medu0", "mage0", "bwgfinal", "smoke0", "leukocyte", "city")) != 11){
    stop("Missing variables in phenotype file, please include: 
         ID, exposure (defined in function), depression, cm1bsex; priorkids; medu0; mage0; bwgfinal;
         smoke0; leukocyte; city")
                                  }
  if(exists("dnam") == F){
    stop("dnam not found 
         Please include DNAm file with ID as rownames and CpG ID as column names")}
  if(exists("exposure") == F){
    stop("exposure not defined - please include which of the exposures will be analysed")}
  if(exists("loci")==F){
    stop("Please load vector of loci to be analyzed")}
  
  print("Formatting data")
  dnam <- dnam[match(pheno$ID, rownames(dnam)),]
  dnam <- dnam[, colnames(dnam) %in% unique(loci)]

  print(paste0("Beginning mediation analyses of ", exposure))
  mediationRes <- lapply(1:length(loci), function(x){
    #tracking progress
    if(x %% 5 == 0 | x ==1){print(paste0("Mediation #", x," of ", length(loci)))}
    #formatting data
    
    #NOTE JONAH - add the array covariate to the following lines if including 
    full <- data.frame(ID = pheno$ID, 
                       depression = pheno$depression, 
                       exposure = pheno[, colnames(pheno)==exposure ],
                       cpg = dnam[,colnames(dnam) == loci[x]],
                       pheno[,c("cm1bsex", "priorkids", "medu0", "mage0", "bwgfinal", "smoke0", 
                                "leukocyte", "city")]
                       )
    #beginning mediation
    mediator <- lm(cpg ~ exposure + cm1bsex + priorkids + medu0 + mage0 + bwgfinal + smoke0 + 
                     leukocyte + city, data = full)
    dv <- lm(depression ~ exposure + cpg  + cm1bsex + priorkids + medu0 + mage0 + bwgfinal + smoke0 + 
                     leukocyte + city, data = full)
    set.seed(88) #so it is reproducible
    results = suppressMessages(mediate(mediator, dv, treat='exposure', mediator='cpg', boot=T))
    
    #obtaining mediation results
    temp <- extract_mediation_summary(results)
    temp <- data.frame(CpG = alspacLoci[x],
                       Adversity = exposure, 
                       Mediation = rownames(temp),
                       temp
                       )
  })
  mediationRes <- do.call(rbind, mediationRes)
  rownames(mediationRes) <- 1:nrow(mediationRes)
  print("Analysis complete - returning results")
  colnames(mediationRes) <- gsub("X95..", "", colnames(mediationRes))
  return(mediationRes)
}

```


*Run mediation and save file*
```{r site-specific mediation}
#notes: we ran a SLCMA of adversity -> depression in ALSPAC to determine the period to analyze. These were the findings: 
  # oneadult = 47 months
  # mompsych = 33 months
  # financial hardship = 61 months 
  # faminst = 57 months
  # parental cruelty= 8 months 
# probably want to use similar periods in FFCWS, rather than rerunning that first piece. 
#also might make sense to do ever/never exposed, which was created earlier
#following is a bit messy and will probably take a couple of hours to run (probably want to use cluster)

#oneadult - since the timepoint is between the two, might run both?
oneadult.3.med <- runMediation(pd,  beta, exposure = "oneadult3", loci = alspacLoci) 
oneadult.5.med <- runMediation(pd,  beta, exposure = "oneadult5", loci = alspacLoci)
oneadult.ever.med <- runMediation(pd,  beta, exposure = "oneadult.ever", loci = alspacLoci)
  
#mompsych - check variable names too
mompsych.3.med <- runMediation(pd,  beta, exposure = "mompsych3", loci = alspacLoci) 
mompsych.ever.med <- runMediation(pd,  beta, exposure = "mompsych.ever", loci = alspacLoci)  

#financial hardship - check variable names too
fh.5.med <- runMediation(pd,  beta, exposure = "fh5", loci = alspacLoci) 
fh.ever.med <- runMediation(pd,  beta, exposure = "fh.ever", loci = alspacLoci)  

#faminst - check variable names too
faminst5.med <- runMediation(pd,  beta, exposure = "faminst5", loci = alspacLoci) 
faminst.ever.med <- runMediation(pd,  beta, exposure = "faminst.ever", loci = alspacLoci)  

#parental cruetly - closest is age 3 in FFCWS
par.3.med <- runMediation(pd,  beta, exposure = "par3", loci = alspacLoci) 
par.ever.med <- runMediation(pd,  beta, exposure = "par.ever", loci = alspacLoci)  


allMediation <- rbind(oneadult.3.med, oneadult.5.med, oneadult.ever.med,
                      mompsych.3.med, mompsych.ever.med,
                      fh.5.med, fh.ever.med,
                      faminst5.med, faminst.ever.med,
                      par.3.med, par.ever.med)

save(allMediation, file = paste0("FFCWS_site_replication_", Sys.Date(), ".Rdata"))

```


